Finiteness in Cubical Type Theory

### 2020-01-26 paper.tex
nearly finished

From both of these types we can derive decidable equality.
\begin{romlemma}
  Any countable type has decidable equality.
\end{romlemma}
\subsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
\subsection{Categories}
\section{Search} \label{search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

@@

All of the finiteness predicates we have seen justify exhaustibility.
We will only prove it once, then, for the weakest:
\begin{romtheorem}
  Kuratowski-finite types are exhaustible.
\end{romtheorem}

Omniscience is stronger than exhaustibility, as we can derive the latter from
the former:
\begin{romlemma} \label{omniscient-is-exhaustible}
  Any omniscient type is exhaustible.
\end{romlemma}
\begin{proof}
  For decidable propositions, we know the following:
  \begin{equation} \label{lem-conversion}
    \Pi (x : A) , P(x) \leftrightarrow \neg \Sigma (x : A) , \neg P(x) 
  \end{equation}
  To derive exhaustibility from omniscience, then, we run the predicate in its
  negated form, and then subsequently negate the result.
  The resulting decision over \(\neg \Sigma(x : A) , \neg P(x)\) can be
  converted into \(\Pi (x : A) , P(x)\).
\end{proof}
We cannot derive, however, that any exhaustible type is omniscient, as we do not
have the inverse of equation~\ref{lem-conversion}:
\begin{equation} \label{choice-conversion}
  \Sigma (x : A) , P(x) \leftrightarrow \neg \Pi (x : A) , \neg P(x)
\end{equation}
Such an equation would allow us to pick a representative element from any type,
which is therefore non-constructive.
In a sense, equation~\ref{lem-conversion} requires a form of LEM on the
proposition (i.e. requires it to be decidable), whereas
equation~\ref{choice-conversion} requires a form of choice.
Those finiteness predicates which are ordered do in fact give us this form of
choice, so the conversion is valid.
As such, all of the ordered finiteness predicates imply omniscience.
Again, we will prove it only for the weakest.
\begin{romtheorem}
  Manifest enumerable types are omniscient.
\end{romtheorem}
Finally, we do have a form of omniscience for prop-valued predicates, as they do
not care about the chosen representative.
\begin{romtheorem}
  Kuratowski finite types are omniscient about prop-valued predicates.
\end{romtheorem}
\subsection{Synthesising Pattern-Matching Proofs}
In particular, they can automate large proofs by analysing every possible case.
In \cite{firsovDependentlyTypedProgramming2015}, the \(\AgdaDatatype{Pauli}\)
group is used as an example.
\ExecuteMetaData[agda/Data/Pauli.tex]{def}
As \(\AgdaDatatype{Pauli}\) has 4 constructors, \(n\)-ary functions on
\(\AgdaDatatype{Pauli}\) may require up to \(4^n\) cases, making even simple
proofs prohibitively verbose.

The alternative is to derive the things we need from \(\mathcal{E}!\) somehow.
As \(\AgdaDatatype{Pauli}\) is a simple finite type, the instance can be defined
in a similar way to those in lemma~\ref{split-enum-bool-top-bot}.
From here we can already derive decidable equality, a function which requires 16
cases if implemented manually.

@@

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{assoc-prf}
\subsection{Synthesising Functions}
Finally, thanks to extensionality provided to us by HoTT, and the computation
properties provided by CuTT, we can derive decidable equality on functions over
finite types.
We can also use functions in our proof search.
Here, for instance, is a automated procedure which finds the
\(\AgdaFunction{not}\) function on \(\mathbf{Bool}\), given a specification.
\ExecuteMetaData[agda/Data/Pauli.tex]{not-spec}

\todo[inline]{Add funding ack:

  This work has been supported by the Science Foundation Ireland under the
  following grant: 13/RC/2D94 to Irish Software Research Centre.
}
\bibliographystyle{splncs04}
\bibliography{bibliography.bib}
\end{document}

>>>>

From both of these types we can derive decidable equality.
\begin{romlemma}
  Any countable type has decidable equality.
\end{romlemma}
\subsection{Closure}
We know that countable infinity is not closed under the exponential (function
arrow), so the only closure we need to prove is \(\Sigma\) to cover all of
what's left.
\begin{romtheorem} \label{split-countability-sigma}
  Split countability is closed under \(\Sigma\).
\end{romtheorem}
This proof does not mirror the proof of \(\Sigma\) closure on finite types,
surprisingly.
A different pattern is needed to pair up the members of each support stream: a
\emph{Cantor} pairing pattern.
Using the simple finite pairing function would diverge, not properly exploring
one of the two lists.
Instead, our function works in two stages.
First, it builds a stream of non-empty lists.
From two support streams \(\mathit{xs}\) and \(\mathit{ys}\), this intermediate
stream \(\mathcal{CV}\) is defined like so:
\begin{equation}
  \mathcal{CV}_i \coloneqq \left[ (\mathit{xs}_j , \mathit{ys}_k) \mid j, k \in \mathbb{N}; j + k = i \right]
\end{equation}
This is then concatenated to give our result.

Finally, while we have lost certain closure proofs by allowing for infinite
types, we also \emph{gain} some: in particular the Kleene star.
\begin{romtheorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{romtheorem}
Again, this proof requires a particular pattern to ensure non divergence.
The pattern here is slightly more complicated, but works on the same basic idea,
building an intermediate stream \(\mathcal{KV}\) of non-empty lists from the
input support stream \(\mathit{xs}\).
\begin{equation}
  \mathcal{KV}_i \coloneqq \left[ \left[ \mathit{xs}_{j - 1} \mid j \in \mathit{js} \right] \mid \mathit{js} \in \mathbf{List}(\mathbb{N}) ; \text{sum}(\mathit{js}) = i ; 0 \notin \mathit{js}  \right]
\end{equation}
Again, this is subsequently flattened.
\section{Search} \label{search}
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{romdefinition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{romdefinition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

@@

All of the finiteness predicates we have seen justify exhaustibility.
We will only prove it once, then, for the weakest:
\begin{romtheorem}
  Kuratowski-finite types are exhaustible.
\end{romtheorem}
Omniscience is stronger than exhaustibility, as we can derive the latter from
the former:
\begin{romlemma} \label{omniscient-is-exhaustible}
  Any omniscient type is exhaustible.
\end{romlemma}
All of the ordered finiteness predicates imply omniscience.
We will again only prove it for the weakest.
\begin{romtheorem}
  Manifest enumerable types are omniscient.
\end{romtheorem}
Finally, we do have a form of omniscience for prop-valued predicates, as they do
not care about the chosen representative.
\begin{romtheorem}
  Kuratowski finite types are omniscient about prop-valued predicates.
\end{romtheorem}
\subsection{Synthesising Pattern-Matching Proofs}
In particular, they can automate large proofs by analysing every possible case.
In \cite{firsovDependentlyTypedProgramming2015}, the \(\AgdaDatatype{Pauli}\)
group is used as an example.
\ExecuteMetaData[agda/Data/Pauli.tex]{def}
As \(\AgdaDatatype{Pauli}\) has 4 constructors, \(n\)-ary functions on
\(\AgdaDatatype{Pauli}\) may require up to \(4^n\) cases, making even simple
proofs prohibitively verbose.

The alternative is to derive the things we need from \(\mathcal{E}!\).
From here we can already derive decidable equality, a function which requires 16
cases if implemented manually.

@@

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{assoc-prf}
\subsection{Synthesising Functions}
Finally, thanks to extensionality provided to us by HoTT, and the computation
properties provided by CuTT, we can derive decidable equality on functions over
finite types.
We can also use functions in our proof search.
Here, for instance, is a automated procedure which finds the
\(\AgdaFunction{not}\) function on \(\mathbf{Bool}\), given a specification.
\ExecuteMetaData[agda/Data/Pauli.tex]{not-spec}
\section{Related Work}
\section*{Acknowledgement}
This work has been supported by the Science Foundation Ireland under the
following grant: 13/RC/2D94 to Irish Software Research Centre.
\bibliographystyle{splncs04}
\bibliography{bibliography.bib}
\end{document}
